Something about Performance, 2D Component of NEMO. (TBC)

### 2015-03-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
Add more text using parallel efficiency metric

\author{J.~Appleyard, H.~Lieu, J.~Holt, R.~W.~Ford, G.~D.~Riley and A.~R.~Porter% <-this % stops a space
\thanks{J.~Appleyard is with NVIDIA.}%
\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
}

@@

As with the serial optimisations, we consider the effect of each of
the OpenMP optimisation steps described in Section~\ref{sec_omp_steps}
for the $256^2$
domain. Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} show the performance of each of the
versions of the code for the Gnu, Intel and Cray compilers,
respectively.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
  socket. The 24-thread runs employed hyperthreading.}
\label{FIG_gnu_omp_256}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_intel}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
  socket. The 24-thread runs employed hyperthreading.}
\label{FIG_intel_omp_256}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_cray}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
  socket. The 24-thread runs employed hyperthreading.}
\label{FIG_cray_omp_256}
\end{figure}

@@

@@

\begin{figure}
\centering
\includegraphics[width=\picwidth]{cpu_cf_gpu}
\caption{Performance of the best OpenMP-parallel version of \psykal
  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation.}
\label{FIG_cpu_cf_gpu}
\end{figure}

>>>>

\author{J.~Appleyard, R.~W.~Ford, J.~Holt, H.~Lieu, A.~R.~Porter and G.~D.~Riley% <-this % stops a space
\thanks{J.~Appleyard is with NVIDIA.}%
\thanks{H.~Lieu and J.~Holt are with the National Oceanography Centre, Liverpool, UK.}%
\thanks{A.~R.~Porter and R.~Ford are with the Science and Technology Facilities Council, UK.}% <-this % stops a space
\thanks{G.~Riley is with the University of Manchester, Manchester, UK.}% <-this % stops a space
}

@@

As with the serial optimisations, we consider the effect of each of
the OpenMP optimisation steps described in Section~\ref{sec_omp_steps}
for the $256^2$ domain. For this we principally use a single Intel,
Ivy Bridge socket which has 12 hardware cores and support for up to 24
threads with hyperthreading.
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} show the performance of each of the
versions of the code on this system for the Gnu, Intel and Cray
compilers, respectively.

In order to quantify the scaling behaviour of the different versions
of nemolite2d with the different compilers/run-time environments, we
also plot the parallel efficiency in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} (dashed lines and open symbols). We define
parallel efficiency (\%), $E(n)$, on a number of threads, $n$, as:
\[
E(n) = 100  \frac{P(n)}{ n P(1)}
\]
where $P(m)$ is the performance of the code on $m$ threads. For a
perfect, linearly-scaling code, $E(n)$ will be 100\%.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_gnu}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Gnu compiler on a single Intel Ivy Bridge
  socket. The corresponding parallel efficiencies are shown using open
  symbols and dashed lines. The 24-thread runs employed hyperthreading
  and the optimal OpenMP schedule is given in parentheses.}
\label{FIG_gnu_omp_256}
\end{figure}

The simplest OpenMP implementation (black lines, circle symbols) fails
to scale well for any of the compilers. For the Intel and Gnu
versions, parallel efficiency is already less than 50\% on just four
threads. The Cray version however does better and is about 45\%
efficient on eight threads.

With the move to a single PARALLEL region, the situation
is greatly improved with all three executables now scaling out to at
least 12 threads with $\sim70$\% efficiency (red lines, square symbols).

In restricting ourselves to a single socket, we are keeping all
threads within a single NUMA region. It is therefore surprising that
implementing a `first-touch' policy has any effect and yet, for the
Gnu- and Intel-compiled binaries, it appears to improve performance
when hyperthreading is employed to run on 24 threads (green lines and
diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
~\ref{FIG_intel_omp_256}). {\bf Can we explain this?}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_intel}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Intel compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_intel_omp_256}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\picwidth]{omp_scaling_256_cray}
\caption{Performance of the OpenMP-parallel version of \psykal
  nemolite2d for the Cray compiler on a single Intel Ivy Bridge
  socket.  The corresponding parallel efficiencies are shown using
  open symbols and dashed lines. The 24-thread runs employed
  hyperthreading and the optimal OpenMP schedule is given in
  parentheses.}
\label{FIG_cray_omp_256}
\end{figure}

The final optimisation step that we found to have any significant
effect is to minimise the amount of thread synchronisation by
introducing the NOWAIT qualifier wherever possible (blue lines and
upwards-triangle symbols). For the Gnu compiler, this improves the
performance of the executable on eight or more threads while, for the
Intel compiler, it only gives an improvement for the 24-thread
case. Moving the SINGLE region before a parallel loop is marginally
beneficial for the Gnu- and Cray-compiled binaries and yet reduces the
performance of the Intel binary (purple lines and right-pointing
triangles).

We have used different scales for the $y$-axes in each of the plots in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} in order to highlight the performance
differences between the code versions with a given compiler. However,
the best performance obtained on twelve threads ({\it i.e.}  without
hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
Intel and Cray compilers, respectively. This serves to emphasise the
democratisation that the introduction of OpenMP has had; for the
serial case the Cray- and Intel-compiled exectables were a factor of
two faster than the Gnu-compiled binary
(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
Gnu-compiled binary is only 6\% slower than that produced by the Intel
compiler and is 13\% {\em faster} that that of the Cray compiler. In
part this situation comes about because of the effect that adding the
OpenMP compiler flag has on the optimisations performed by the
compiler.  In particular, the Cray compiler no-longer vectorises the
key Momentum section, despite the directive added during the serial
optimisation work.

A further complication is the choice of scheduling of the OpenMP
threads.  We have investigated the performance of each of the
executables (and thus the associated OpenMP run-time library) with the
standard OpenMP {\it static}, {\it dynamic} and {\it guided}
scheduling policies~\cite{openmp_standard}. For the Intel
compiler/run-time, static loop scheduling was found to be best for all
versions apart from that where we have attempted to amortize the cost
of the SINGLE section. This is to be expected since that strategy
requires some form of dynamic loop scheduling in order to reduce the
load imbalance introduced by the SINGLE section.

In contrast, some form of dynamic scheduling gave a performance
improvement with the Gnu compiler/run-time even for the `first-touch'
version of the code. This is despite the fact that this version
contains (implicit) thread synchronisation after every parallel loop.
For the Cray compiler/run-time, some form of dynamic scheduling became
optimal once inter-thread synchronisation was reduced using the NOWAIT
qualifiers.

@@

The introduction of OpenACC directives proved to be simpler than
producing the OpenMP version of nemolite2d {\bf - Jeremy to clarify
  whether this is true}.  Consequently we only show the performance of
{\bf xxx} versions of the code in Figure~\ref{FIG_gpu_opt_stages}.

@@

In Figure~\ref{FIG_cpu_cf_gpu} we compare the absolute performance of
the OpenMP and OpenACC implementations of nemolite2d across the range
of problem sizes considered. The OpenMP figures are the maximum
performance obtained from a whole Intel Ivy Bridge socket by any
version of the code on any number of threads for a given
compiler/run-time. For the smallest domain size ($64^2$) the OpenMP
version significantly outperforms the GPU because there is
insufficient parallelism to fully utilise the GPU and one time-step
takes only 80~$\mu$s. The execution of a single time-step is then
dominated by the time taken to launch a kernel on the GPU rather than
the execution of the kernel itself.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{cpu_cf_gpu}
\caption{Performance of the best OpenMP-parallel version of \psykal
  nemolite2d (on a single Intel Ivy Bridge socket) compared with the best GPU implementation. {\bf This should really be best OpenACC version as that's what PSyclone can support.}}
\label{FIG_cpu_cf_gpu}
\end{figure}

Once the problem size is increased to $128^2$, a single time-step
takes roughly 200~$\mu$s and only the Intel-compiled OpenMP version is
comparable in performance to the OpenACC version. For all of the
larger problem sizes plotted in Figure~\ref{FIG_cpu_cf_gpu} the GPU
version is considerably faster than the CPU.  For problem sizes of
$1024^2$ and greater, the 30MB cache of the Ivy Bridge CPU is
exhausted and performance becomes limited by the bandwidth to main
memory. At this stage the OpenACC version of the code on the GPU is
some {\bf 3.7 (this is the factor for the CUDA version!)} times faster
than the OpenMP version on the CPU. {\bf Can we relate this to the
  difference in peak memory b/w between the GPU and CPU?}
